---
title: "Parallel Differential Evolution with RcppDEpar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel Differential Evolution with RcppDEpar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(RcppDEpar)
```

This vignette exercises `RcppDEpar` the way the scripts in `demo/` do
(`DEoptim.R`, `small.R`, `large.R`, `compiled.R`, `parallel_breeding.R`), but
trimmed to sizes that build quickly, and with the focus on the feature that
sets this fork apart from upstream `RcppDE`: the **population-parallel**
evaluation path (`control$nthreads`).

## 1. Basic usage

`DEoptim()` has the same interface as `DEoptim::DEoptim()` (see
`demo/DEoptim.R`): minimize a function over a box `[lower, upper]`.

```{r basic}
Rosenbrock <- function(x) {
  x1 <- x[1]; x2 <- x[2]
  100 * (x2 - x1 * x1)^2 + (1 - x1)^2
}

set.seed(42)
fit <- DEoptim(Rosenbrock, lower = rep(-10, 2), upper = rep(10, 2),
               control = DEoptim.control(NP = 50, itermax = 100, trace = FALSE))
fit$optim$bestmem
fit$optim$bestval
```

## 2. Two ways to run an objective

`RcppDEpar` accepts either an **R function** (a closure, evaluated on the
main thread through R's C API) or a **compiled function pointer** wrapped in
an external pointer (evaluated in C++, with no R-level call overhead). The
package ships example compiled objectives in `src/exampleFunctions.cpp`,
exposed through `putFunPtrInXPtr()`/`putFunPtrInXPtrRaw()` (used by
`demo/compiled.R` and `demo/parallel_breeding.R`).

```{r compiled-vs-r}
Rastrigin <- function(x) sum(x + 2 - 10 * cos(2 * pi * x)) + 20

set.seed(42)
fitR <- DEoptim(Rastrigin, lower = rep(-25, 5), upper = rep(25, 5),
                control = DEoptim.control(NP = 50, itermax = 100, trace = FALSE))

xptr <- RcppDEpar:::putFunPtrInXPtr("rastrigin")
set.seed(42)
fitCpp <- DEoptim(xptr, lower = rep(-25, 5), upper = rep(25, 5),
                   control = DEoptim.control(NP = 50, itermax = 100, trace = FALSE))

all.equal(fitR$optim$bestval, fitCpp$optim$bestval)
```

Both objectives above run on `devol()`, the sequential engine — one
individual evaluated after another. This matches `RcppDE` upstream.

## 3. The population-parallel path

`RcppDEpar` adds a second engine, `devol_parallel()`, that evaluates an
entire generation's population across `control$nthreads` worker threads
(via `RcppParallel`) instead of one individual at a time. Two conditions
must both hold for a `DEoptim()` call to use it:

1. The objective must be a **raw, thread-safe compiled function pointer**
   created with `putFunPtrInXPtrRaw()` — not an R closure, and not a plain
   `putFunPtrInXPtr()` pointer (different, non-thread-safe calling
   convention). This is checked in `src/deoptim.cpp`'s
   `isRawCompiledObjective()`.
2. `control$nthreads` is set (`> 1` to actually spread work across
   threads; `nthreads = 1` still routes through `devol_parallel()`, just
   on a single worker, since a raw pointer can *only* be evaluated that
   way).

An R closure with `nthreads > 1` will **not** go parallel — it silently
runs on the ordinary sequential `devol()` engine, because only the raw
compiled calling convention is safe to call from worker threads.

The example objective used for this is `breedingPenaltyRaw()`
(`src/exampleFunctions.cpp`): a deliberately branchy, non-vectorizable
per-candidate score (parent-usage cap + local coancestry penalty) standing
in for a breeding-selection objective — see `demo/parallel_breeding.R` for
the full-scale benchmark this vignette is a scaled-down version of.

```{r raw-xptr}
xptrRaw <- RcppDEpar:::putFunPtrInXPtrRaw("breedingPenaltyRaw")
```

### Correctness: same seed, same result, any thread count

Mutation/crossover randomness in the parallel path is seeded per
`(generation, individual index)`, not per thread (see the `ThreadRng` /
`SplitMix64` generator in `src/devol.cpp`), so the result is reproducible
regardless of how many threads run the population loop:

```{r reproducible, warning = FALSE}
## warning = FALSE: NP is deliberately far below 10 * length(lower) here, so
## DEoptim()'s standing advice about population size is expected noise
n  <- 3000
NP <- 500
lower <- rep(0, n); upper <- rep(1, n)

runWith <- function(nthreads) {
  set.seed(42)
  ctrl <- DEoptim.control(NP = NP, itermax = 15, trace = FALSE, nthreads = nthreads)
  DEoptim(xptrRaw, lower, upper, control = ctrl)$optim$bestval
}

bestvals <- sapply(c(1, 2), runWith)
bestvals
stopifnot(abs(diff(bestvals)) < 1e-8)
```

### Benchmark: scaling the genetic-selection objective across 1, 2, 4 and 8 threads

Population-parallel evaluation pays off when the *per-individual*
objective is expensive and the population is large enough to amortize
`RcppParallel`'s scheduling overhead — the `demo/parallel_breeding.R`
regime (`n = 100000`, `NP = 20000`) makes that obvious. The problem below
is larger than the reproducibility check above so that the timings are not
pure noise, but still small enough to build in a few seconds; it is a
scaled-down stand-in, not a definitive benchmark.

Each configuration is run three times and the **minimum** elapsed time is
reported, which is the least noise-prone estimator for this kind of
measurement. Thread counts above the number of available cores are
dropped, so a smaller machine simply shows fewer rows.

```{r benchmark-setup}
nBench  <- 20000    # candidates per solution vector
NPBench <- 10000     # population size
itBench <- 50     # generations
lowerB <- rep(0, nBench); upperB <- rep(1, nBench)

## Respect the 2-core limit R CMD check imposes when _R_CHECK_LIMIT_CORES_ is set
maxCores <- if (nzchar(Sys.getenv("_R_CHECK_LIMIT_CORES_"))) 2L else parallel::detectCores()
threadCounts <- c(1, 2, 4, 8)
threadCounts <- threadCounts[threadCounts <= maxCores]
threadCounts
```

```{r benchmark, warning = FALSE}
runOnce <- function(nthreads) {
  gc()
  set.seed(42)
  ctrl <- DEoptim.control(NP = NPBench, itermax = itBench, trace = FALSE,
                          nthreads = nthreads)
  elapsed <- system.time(
    fit <- DEoptim(xptrRaw, lowerB, upperB, control = ctrl))[["elapsed"]]
  list(seconds = elapsed, bestval = fit$optim$bestval)
}

runBenchmark <- function(nthreads, reps = 3) {
  runs <- lapply(seq_len(reps), function(i) runOnce(nthreads))
  data.frame(nthreads = nthreads,
             seconds  = min(vapply(runs, `[[`, numeric(1), "seconds")),
             bestval  = runs[[1]]$bestval)
}

res <- do.call(rbind, lapply(threadCounts, runBenchmark))
res$speedup    <- res$seconds[1] / res$seconds
res$efficiency <- res$speedup / res$nthreads
res
```

The optimum found is identical at every thread count — the same property
checked in the previous section, now across the full sweep:

```{r benchmark-identical}
stopifnot(all(abs(res$bestval - res$bestval[1]) < 1e-8))
```

```{r benchmark-plot, fig.width = 6, fig.height = 4, fig.alt = "Speedup relative to one thread against thread count, with the ideal linear-scaling line for reference"}
plot(res$nthreads, res$speedup, type = "b", pch = 19, log = "x",
     xaxt = "n", ylim = range(0, 1, res$speedup, na.rm = TRUE),
     xlab = "threads", ylab = "speedup vs. 1 thread",
     main = "Genetic-selection objective: parallel scaling")
axis(1, at = res$nthreads, labels = res$nthreads)
abline(h = 1, lty = 3)                                   # break-even
lines(res$nthreads, res$nthreads, lty = 2, col = "grey") # ideal linear scaling
legend("topleft", bty = "n", lty = c(1, 3, 2), pch = c(19, NA, NA),
       col = c("black", "black", "grey"),
       legend = c("measured", "break-even", "ideal"))
```

#### Reading the table

`speedup` above 1 means the parallel engine beat single-threaded
execution; `efficiency` is that speedup divided by the thread count, so
1.0 would be perfect linear scaling and anything near 0 means the threads
are mostly getting in each other's way.

The generation loop is a hard barrier: each generation dispatches the
whole population through `RcppParallel::parallelFor()` and then joins
before the next one can start, so there is a fixed cost paid `itermax`
times regardless of how much work each individual actually represents.
Here that cost exceeds the per-generation work, and adding threads only
claws back part of what the dispatch itself costs. (The `RcppParallel`
TinyThread backend creates and joins worker threads on every
`parallelFor()` call rather than keeping a pool, which is the leading
suspect — but the non-monotonic 2 → 8 curve above is not fully explained
by that alone, so treat it as unconfirmed.)

The practical guidance is therefore: **measure on your own problem before
turning threads on.** Parallelism starts to win when per-generation work
grows — a more expensive objective, a larger `n`, or a larger `NP` —
because the fixed per-generation dispatch cost stays the same while the
work it is amortized over grows. Increase `nBench`, `NPBench` or
`itBench` above, or run
`demo("parallel_breeding", package = "RcppDEpar")` for a regime sized to
show it.

## 4. Sanity check against base `DEoptim`

As in `demo/small.R`, the sequential engine can be compared against the
reference `DEoptim` package on a simple problem (skipped here if `DEoptim`
isn't installed):

```{r vs-DEoptim, eval = requireNamespace("DEoptim", quietly = TRUE)}
maxIt <- 100

set.seed(42)
valBasic <- DEoptim::DEoptim(Rastrigin, lower = rep(-25, 5), upper = rep(25, 5),
                              control = list(NP = 50, itermax = maxIt, trace = FALSE))
set.seed(42)
valCpp <- DEoptim(Rastrigin, lower = rep(-25, 5), upper = rep(25, 5),
                   control = DEoptim.control(NP = 50, itermax = maxIt, trace = FALSE))

c(DEoptim = valBasic$optim$bestval, RcppDEpar = valCpp$optim$bestval)

## same solution quality, not the same trajectory -- see below
all.equal(valBasic$optim$bestval, valCpp$optim$bestval, tolerance = 0.05)
```

Note that this is a **quality** comparison, not a bit-for-bit one. `RcppDE`
— and therefore this fork — is a port of `DEoptim` **2.0-7**, while the
`DEoptim` package on CRAN has moved on:

```{r DEoptim-version, eval = requireNamespace("DEoptim", quietly = TRUE)}
packageVersion("DEoptim")
```

From the same seed the two implementations consume R's RNG stream
differently — they already disagree at the *first* recorded generation,
before any mutation has happened — so their search trajectories diverge
and only the quality of the final optimum is meaningfully comparable. Do
not read a small difference here as a bug in either package.

## Summary

* Plain R-function and `putFunPtrInXPtr()` objectives always run
  sequentially through `devol()`, matching upstream `RcppDE`.
* `control$nthreads > 1` only takes effect for objectives created with
  `putFunPtrInXPtrRaw()`; these dispatch to the population-parallel
  `devol_parallel()` engine.
* Results are bit-for-bit reproducible across thread counts because RNG
  streams are seeded per `(generation, individual)`, not per thread.
* Parallelism is not free and not automatically a win: each generation is
  a barrier, so a fixed dispatch cost is paid `itermax` times. At the size
  used here the 1/2/4/8-thread sweep comes out *below* break-even on the
  development machine. Speedup only arrives once per-generation work
  (objective cost, `n`, `NP`) is large enough to amortize that — measure
  before enabling threads, and see `demo/parallel_breeding.R` for a
  regime sized to show it.
