---
title: "Parallel Differential Evolution with RcppDEpar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel Differential Evolution with RcppDEpar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(RcppDEpar)
```

This vignette exercises `RcppDEpar` the way `vignettes/parallel_breeding.R`
does, but trimmed to sizes that build quickly, and with the focus on the
feature that sets this fork apart from upstream `RcppDE`: the
**population-parallel** evaluation path (`control$nthreads`).

## 1. Basic usage

`DEoptim()` has the same interface as `DEoptim::DEoptim()` (see
`demo/DEoptim.R`): minimize a function over a box `[lower, upper]`.

```{r basic}
Rosenbrock <- function(x) {
  x1 <- x[1]; x2 <- x[2]
  100 * (x2 - x1 * x1)^2 + (1 - x1)^2
}

set.seed(42)
fit <- DEoptim(Rosenbrock, lower = rep(-10, 2), upper = rep(10, 2),
               control = DEoptim.control(NP = 50, itermax = 100, trace = FALSE))
fit$optim$bestmem
fit$optim$bestval
```

## 2. Two ways to run an objective

`RcppDEpar` accepts either an **R function** (a closure, evaluated on the
main thread through R's C API) or a **compiled function pointer** wrapped in
an external pointer (evaluated in C++, with no R-level call overhead). The
package ships example compiled objectives in `src/exampleFunctions.cpp`,
exposed through `putFunPtrInXPtr()`/`putFunPtrInXPtrRaw()` (used by
`vignettes/parallel_breeding.R`).

```{r compiled-vs-r}
Rastrigin <- function(x) sum(x + 2 - 10 * cos(2 * pi * x)) + 20

set.seed(42)
fitR <- DEoptim(Rastrigin, lower = rep(-25, 5), upper = rep(25, 5),
                control = DEoptim.control(NP = 50, itermax = 100, trace = FALSE))

xptr <- RcppDEpar:::putFunPtrInXPtr("rastrigin")
set.seed(42)
fitCpp <- DEoptim(xptr, lower = rep(-25, 5), upper = rep(25, 5),
                   control = DEoptim.control(NP = 50, itermax = 100, trace = FALSE))

all.equal(fitR$optim$bestval, fitCpp$optim$bestval)
```

Both objectives above run on `devol()`, the sequential engine — one
individual evaluated after another. This matches `RcppDE` upstream.

## 3. The population-parallel path

`RcppDEpar` adds a second engine, `devol_parallel()`, that evaluates an
entire generation's population across `control$nthreads` worker threads
(via `RcppParallel`) instead of one individual at a time. Two conditions
must both hold for a `DEoptim()` call to use it:

1. The objective must be a **raw, thread-safe compiled function pointer**
   created with `putFunPtrInXPtrRaw()` — not an R closure, and not a plain
   `putFunPtrInXPtr()` pointer (different, non-thread-safe calling
   convention). This is checked in `src/deoptim.cpp`'s
   `isRawCompiledObjective()`.
2. `control$nthreads` is set (`> 1` to actually spread work across
   threads; `nthreads = 1` still routes through `devol_parallel()`, just
   on a single worker, since a raw pointer can *only* be evaluated that
   way).

`nthreads` is authoritative in both directions: `DEoptim()` calls
`RcppParallel::setThreadOptions()` on every raw-objective call, including
`nthreads = 1`, and restores the previous process-wide setting on exit. That
matters for benchmarking — without it, `nthreads = 1` would inherit
`RcppParallel`'s default (every core on the machine), so the "serial"
baseline would itself be a fully parallel run and every reported speedup
would collapse to roughly 1.

An R closure with `nthreads > 1` will **not** go parallel — it silently
runs on the ordinary sequential `devol()` engine, because only the raw
compiled calling convention is safe to call from worker threads.

The example objective used for this is `breedingPenaltyRaw()`
(`src/exampleFunctions.cpp`): a deliberately branchy, non-vectorizable
per-candidate score (parent-usage cap + local coancestry penalty) standing
in for a breeding-selection objective — see `demo/parallel_breeding.R` for
the full-scale benchmark this vignette is a scaled-down version of.

```{r raw-xptr}
xptrRaw <- RcppDEpar:::putFunPtrInXPtrRaw("breedingPenaltyRaw")
```

### Correctness: same seed, same result, any thread count

Mutation/crossover randomness in the parallel path is seeded per
`(generation, individual index)`, not per thread (see the `ThreadRng` /
`SplitMix64` generator in `src/devol.cpp`), so the result is reproducible
regardless of how many threads run the population loop:

```{r reproducible, warning = FALSE}
## warning = FALSE: NP is deliberately far below 10 * length(lower) here, so
## DEoptim()'s standing advice about population size is expected noise
n  <- 10000
NP <- 2000
lower <- rep(0, n); upper <- rep(1, n)

runWith <- function(nthreads) {
  set.seed(42)
  ctrl <- DEoptim.control(NP = NP, itermax = 15, trace = FALSE, nthreads = nthreads)
  DEoptim(xptrRaw, lower, upper, control = ctrl)$optim$bestval
}

bestvals <- sapply(c(1, 2), runWith)
bestvals
stopifnot(abs(diff(bestvals)) < 1e-8)
```

### Benchmark: scaling the genetic-selection objective across thread counts

Population-parallel evaluation pays off when the *per-individual* objective
is expensive and the population is large enough to amortize `RcppParallel`'s
scheduling overhead — the `vignettes/parallel_breeding.R` regime
(`n = 10000`, `NP = 2000`) makes that obvious. The problem below is larger
than the reproducibility check above so that the timings are not pure noise,
but still small enough to build in a few seconds; it is a scaled-down
stand-in, not a definitive benchmark.

Each configuration is run three times and the **minimum** elapsed time is
reported, which is the least noise-prone estimator for this kind of
measurement.

Thread counts above the number of available cores are dropped, so a smaller
machine simply shows fewer rows. This filter is not cosmetic: asking for more
workers than there are cores *oversubscribes* the machine — the OS
time-slices the workers onto the same cores, and since each generation is a
barrier, the whole population waits for whichever chunk got descheduled. On
the 18-core development machine, 24 threads measured **slower** than 16
(speedup 6.5 vs 7.3). Never set `nthreads` above the core count.

`detectCores(logical = FALSE)` (physical cores) is the right ceiling rather
than the logical count: with hyper-threading, two logical siblings share one
core's execution units, and this loop is memory- rather than branch-bound, so
the second sibling adds contention more than throughput.

```{r benchmark-setup}
nBench  <- 1000   # candidates per solution vector
NPBench <- 1000   # population size
itBench <- 50     # generations
lowerB <- rep(0, nBench); upperB <- rep(1, nBench)

## Respect the 2-core limit R CMD check imposes when _R_CHECK_LIMIT_CORES_ is set
maxCores <- if (nzchar(Sys.getenv("_R_CHECK_LIMIT_CORES_"))) {
  2L
} else {
  cores <- parallel::detectCores(logical = FALSE)
  if (is.na(cores)) 1L else cores          # detectCores() returns NA on some platforms
}
threadCounts <- c(1, 2, 4, 8, 12, 16, 24)
threadCounts <- threadCounts[threadCounts <= maxCores]
threadCounts
```

```{r benchmark, warning = FALSE}
runOnce <- function(nthreads) {
  gc()
  set.seed(42)
  ctrl <- DEoptim.control(NP = NPBench, itermax = itBench, trace = FALSE,
                          nthreads = nthreads)
  elapsed <- system.time(
    fit <- DEoptim(xptrRaw, lowerB, upperB, control = ctrl))[["elapsed"]]
  list(seconds = elapsed, bestval = fit$optim$bestval)
}

runBenchmark <- function(nthreads, reps = 3) {
  runs <- lapply(seq_len(reps), function(i) runOnce(nthreads))
  data.frame(nthreads = nthreads,
             seconds  = min(vapply(runs, `[[`, numeric(1), "seconds")),
             bestval  = runs[[1]]$bestval)
}

res <- do.call(rbind, lapply(threadCounts, runBenchmark))
res$speedup    <- res$seconds[1] / res$seconds
res$efficiency <- res$speedup / res$nthreads
res
```

The optimum found is identical at every thread count — the same property
checked in the previous section, now across the full sweep:

```{r benchmark-identical}
stopifnot(all(abs(res$bestval - res$bestval[1]) < 1e-8))
```

```{r benchmark-plot, fig.width = 6, fig.height = 5, fig.alt = "Speedup relative to one thread against thread count, with the ideal linear-scaling line for reference", echo =FALSE}
plot(res$nthreads, res$speedup, type = "b", pch = 19, log = "x",
     xaxt = "n", ylim = range(0, 1, res$speedup, na.rm = TRUE),
     xlab = "threads", ylab = "speedup vs. 1 thread",
     main = "Genetic-selection objective: parallel scaling")
axis(1, at = res$nthreads, labels = res$nthreads)
abline(h = 1, lty = 3)                                   # break-even
lines(res$nthreads, res$nthreads, lty = 2, col = "grey") # ideal linear scaling
legend("topleft", bty = "n", lty = c(1, 3, 2), pch = c(19, NA, NA),
       col = c("black", "black", "grey"),
       legend = c("measured", "break-even", "ideal"))
```

#### Reading the table

`speedup` above 1 means the parallel engine beat single-threaded
execution; `efficiency` is that speedup divided by the thread count, so
1.0 would be perfect linear scaling and anything near 0 means the threads
are mostly getting in each other's way.

On the 18-core development machine this sweep gives 1.88x at 2 threads
(efficiency 0.94), 3.37x at 4, 5.51x at 8 and 7.29x at 16 — a real win that
flattens rather than reverses.

Efficiency falls off with thread count for two reasons, and it is worth
separating them because only one of them is a property of the algorithm.

**Amdahl, not scheduling overhead.** Part of each `DEoptim()` call is
irreducibly serial: allocating and zeroing the `D x NP` population matrices,
the initialization loop in `devol_parallel()` (which evaluates all `NP`
starting members on the main thread), and transposing the results back to R.
Sweeping `itermax` on the larger `parallel_breeding.R` problem separates the
fixed cost from the per-generation cost cleanly:

| `itermax` | 1 thread | 16 threads | speedup |
|---|---|---|---|
| 1 | 0.288 s | 0.198 s | 1.42 |
| 5 | 0.675 s | 0.250 s | 2.79 |
| 10 | 1.169 s | 0.299 s | 4.65 |
| 30 | 2.856 s | 0.348 s | 8.09 |
| 60 | 5.527 s | 0.532 s | 10.47 |

Both series are linear in `itermax` with the *same* intercept, about
0.195 s — that is the serial fixed cost. The slopes are 0.0888 s/generation
on one thread against 0.0057 s/generation on sixteen, i.e. the population
loop itself scales at **15.7x on 16 threads**, near-perfect. The whole-run
speedup is lower only because the fixed 0.195 s is a growing share of the
total as the parallel part shrinks; it rises steadily with `itermax` exactly
as Amdahl's law predicts.

**Heterogeneous cores.** `RcppParallel`'s TinyThread backend partitions the
population statically and has no work-stealing, so the generation barrier
waits for the slowest chunk. On a CPU with performance and efficiency cores
(Apple silicon, Intel P/E hybrids), the chunks landing on efficiency cores
take several times longer and set the pace for everyone. That is why 18
threads is no better than 16 on the development machine, and why the last few
cores are usually not worth requesting.

The practical guidance is therefore: **measure on your own problem, and cap
`nthreads` at the physical core count.** Parallelism wins more as
per-generation work grows — a more expensive objective, a larger `n`, or a
larger `NP` — because the serial fixed cost stays put while the work it is
amortized over grows. Increase `nBench`, `NPBench` or `itBench` above, or run
`vignettes/parallel_breeding.R` for a regime sized to show it.

## 4. Sanity check against base `DEoptim`

As in `demo/small.R`, the sequential engine can be compared against the
reference `DEoptim` package on a simple problem (skipped here if `DEoptim`
isn't installed):

```{r vs-DEoptim, eval = requireNamespace("DEoptim", quietly = TRUE)}
maxIt <- 100

set.seed(42)
valBasic <- DEoptim::DEoptim(Rastrigin, lower = rep(-25, 5), upper = rep(25, 5),
                              control = list(NP = 50, itermax = maxIt, trace = FALSE))
set.seed(42)
valCpp <- DEoptim(Rastrigin, lower = rep(-25, 5), upper = rep(25, 5),
                   control = DEoptim.control(NP = 50, itermax = maxIt, trace = FALSE))

c(DEoptim = valBasic$optim$bestval, RcppDEpar = valCpp$optim$bestval)

## same solution quality, not the same trajectory -- see below
all.equal(valBasic$optim$bestval, valCpp$optim$bestval, tolerance = 0.05)
```

Note that this is a **quality** comparison, not a bit-for-bit one. `RcppDE`
— and therefore this fork — is a port of `DEoptim` **2.0-7**, while the
`DEoptim` package on CRAN has moved on:

```{r DEoptim-version, eval = requireNamespace("DEoptim", quietly = TRUE)}
packageVersion("DEoptim")
```

From the same seed the two implementations consume R's RNG stream
differently — they already disagree at the *first* recorded generation,
before any mutation has happened — so their search trajectories diverge
and only the quality of the final optimum is meaningfully comparable. Do
not read a small difference here as a bug in either package.

## Summary

* Plain R-function and `putFunPtrInXPtr()` objectives always run
  sequentially through `devol()`, matching upstream `RcppDE`.
* `control$nthreads > 1` only takes effect for objectives created with
  `putFunPtrInXPtrRaw()`; these dispatch to the population-parallel
  `devol_parallel()` engine.
* Results are bit-for-bit reproducible across thread counts because RNG
  streams are seeded per `(generation, individual)`, not per thread.
* The population loop itself scales close to linearly (measured 15.7x on 16
  threads). Whole-call speedup is lower because each `DEoptim()` call carries
  a fixed serial cost — matrix allocation, the main-thread initialization
  loop, transposing results back to R — that does not shrink with threads.
  Speedup therefore grows with `itermax`, `n` and `NP`; see
  `vignettes/parallel_breeding.R` for a regime sized to show it.
* Never set `nthreads` above the physical core count. Oversubscription is a
  net loss (24 threads on 18 cores measured slower than 16), and because the
  scheduler partitions statically with no work-stealing, efficiency cores set
  the pace for the whole generation on hybrid CPUs.
